---
title: "README"
author: "Caleb Rother"
date: "2025-06-27"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Sourcing files with relative paths makes the project portable.
# This assumes the .R files are in the same directory as the .Rmd file.
source("paired_comparison.R")
source("group_comparisons.R")
source("anova_assumptions.R")
```

## Web Interface

This project includes a lightweight browser interface in `index.html` that
leverages WebR. You can import a CSV file (using the button or by dragging and
dropping it onto the spreadsheet) or manually enter data into the embedded
spreadsheet. After editing, the table can be exported back to a CSV file.

# **Function: paired_comparison()**

## Purpose

Performs a paired t-test or its non-parametric equivalent (Wilcoxon
Signed-Rank Test) to determine if there is a significant change between
two paired measurements. It automatically generates a plot and prints
the statistical results.

## When to Use

This function is designed exclusively for paired or dependent data. The
classic use case is a "before-and-after" study where two measurements
are taken from the same subject or a matched pair.

**Example**: Measuring a patient's biomarker level before and after a
specific treatment.

**Warning**: Do not use this for comparing two independent groups (e.g.,
a treatment group vs. a separate control group).

## Arguments

| Argument | Description | Required? | Default Value |
|:-----------------|:-----------------|:------------------:|:-----------------|
| `data` | The input data frame. | Yes |  |
| `before_col` | The unquoted column name for the "before" measurement. | Yes |  |
| `after_col` | The unquoted column name for the "after" measurement. | Yes |  |
| `parametric` | Logical. `TRUE` for a Paired t-Test, `FALSE` for Wilcoxon Signed-Rank. | No | `FALSE` |
| `plot_title` | A custom character string for the plot's main title. | No | Auto-generated |
| `xlab` | A custom character string for the x-axis label. | No | `"Time Point"` |
| `ylab` | A custom character string for the y-axis label. | No | `"Value"` |
| `before_label` | A custom string to display for the "before" group on the x-axis. | No | Uses the column name |
| `after_label` | A custom string to display for the "after" group on the x-axis. | No | Uses the column name |
| `show_paired_lines` | Logical. If `TRUE`, shows the grey lines connecting paired points. | No | `TRUE` |
| `before_color` | A string for the "before" group's color (e.g., `"blue"`, `"#FF6347"`). | No | ggplot2 default |
| `after_color` | A string for the "after" group's color. | No | ggplot2 default |

```{r create-paired-data}
# Create Sample Paired Data
set.seed(42)
n_subjects <- 80
before_after_data <- data.frame(
  patient_id = paste0("P", 1:n_subjects),
  biomarker_baseline = rnorm(n_subjects, mean = 150, sd = 20)
)
before_after_data$biomarker_followup <- before_after_data$biomarker_baseline + rnorm(n_subjects, mean = -15, sd = 10)
```

```{r paired-example-1}
paired_comparison(
  data = before_after_data,
  before_col = biomarker_baseline,
  after_col = biomarker_followup
)
```

```{r paired-example-2}
paired_comparison(
  data = before_after_data,
  before_col = biomarker_baseline,
  after_col = biomarker_followup,
  parametric = TRUE,
  plot_title = "Biomarker Levels After 6-Month Treatment",
  ylab = "Biomarker Concentration (U/L)",
  xlab = "Study Phase",
  before_label = "Baseline Visit",
  after_label = "6-Month Follow-up",
  show_paired_lines = FALSE,
  before_color = "#1B9E77",
  after_color = "#7570B3"
)
```

<br>

<hr>

<br>

# **Function: `check_anova_assumptions()`**

## Purpose

A diagnostic tool to check the statistical assumptions required to run a
parametric ANOVA. It performs formal tests for normality of residuals
and homogeneity of variances, provides visual plots, and gives a clear
recommendation on whether to use a parametric or non-parametric test.

## When to Use

Use this *before* running `group_comparison` with `parametric = TRUE`.
It helps you justify your choice of test. It works for one-way and
two-way models.

## Arguments

| Argument | Description | Required? | Default |
|:------------|:----------------------------------|:-----------:|:------------|
| `data` | Input data frame (tibble or data.frame). | Yes |  |
| `formula` | An R formula specifying the model (e.g., `Response ~ Group1 * Group2`). | Yes |  |

## Example

```{r anova-check-example}
# Create sample data for a one-way comparison
set.seed(123)
df_check <- data.frame(
  Group    = factor(rep(c("A", "B", "C"), each = 20)),
  Outcome  = c(rnorm(20, 5, 1), rnorm(20, 6, 1.5), rnorm(20, 7, 2.5)) # Variances might be unequal
)

# Check the assumptions for a one-way ANOVA model
check_anova_assumptions(
  data = df_check,
  formula = Outcome ~ Group
)
```

<br>

<hr>

<br>

# **Function: `group_comparison()`**

## Purpose

Performs one-way or two-way group comparisons using either parametric
tests (ANOVA + Tukey HSD) or non-parametric tests (Kruskal-Wallis +
Dunn’s test). Automatically generates a boxplot with jittered points,
adds pairwise significance annotations, and prints the statistical
results.

## When to Use

-   **One-Way Comparison (single factor):** comparing a numeric outcome
    across two or more independent groups.
-   **Two-Way Comparison (factorial):** assessing the effect of two
    grouping factors (and their interaction) on a numeric outcome.

## Arguments

| Argument | Description | Required? | Default |
|:-------------|:-----------------------------|:------------:|:-------------|
| `data` | Input data frame (tibble or data.frame). | Yes |  |
| `response_col` | Unquoted name of the numeric response column. | Yes |  |
| `group1_col` | Unquoted name of the primary grouping factor. | Yes |  |
| `group2_col` | Unquoted name of the secondary grouping factor (activates two-way). | No | `NULL` |
| `parametric` | Logical: `TRUE` for ANOVA, `FALSE` for non-parametric. | No | `FALSE` |
| `p_label_style` | How to display p-values: `"letters"`, `"p.value"`, or `"both"`. | No | `"letters"` |
| `show_test_name` | Logical: include the main test name in the subtitle. | No | `TRUE` |
| `plot_title` | Character: custom main title for the plot. | No | auto-generated |
| `xlab` | Character: custom label for the x-axis. | No | grouping column name |
| `ylab` | Character: custom label for the y-axis. | No | response column name |
| `palette` | `ggpubr` palette name or vector of colours. | No | `"npg"` |
| `label_size` | Numeric: font size for significance labels. | No | `3.8` |
| `hide_ns` | Logical: if `TRUE`, suppress "ns" annotations. | No | `FALSE` |

## Examples

### 1. One-Way, Non-Parametric (default)

```{r one-way-default}
set.seed(123)
df1 <- data.frame(
  Group    = factor(rep(c("A", "B", "C"), each = 20)),
  Outcome  = c(rnorm(20,5,1), rnorm(20,6,1), rnorm(20,7,1))
)

group_comparison(
  data         = df1,
  response_col = Outcome,
  group1_col   = Group
)
```

### 2. One-Way, Parametric + Both Labels + Custom Palette

```{r one-way-param-both}
group_comparison(
  data          = df1,
  response_col  = Outcome,
  group1_col    = Group,
  parametric    = TRUE,
  p_label_style = "both",
  palette       = c("#D55E00","#0072B2","#009E73")
)
```

### 3. Two-Way ANOVA + Tukey HSD

```{r two-way-parametric}
set.seed(456)
df2 <- expand.grid(
  Genotype = factor(c("WT","Mut")),
  Treatment= factor(c("Placebo","Drug")),
  n = 1:15
)
df2$Response <- rnorm(
  n = nrow(df2),
  mean = with(df2, ifelse(Genotype == "WT", 5, 7) + ifelse(Treatment == "Drug", 2, 0)),
  sd   = 1
)


# CORRECTED: Updated syntax
group_comparison(
  data          = df2,
  response_col  = Response,
  group1_col    = Treatment,
  group2_col    = Genotype,
  parametric    = TRUE
)
```
